A thermodynamic view of dusty protoplanetary disks

### 2017-08-04 linear.tex
misc minor edits
removed higher order correction section for submission, misc edits

\subsubsection{Orbital frequency} 
From Eq. \ref{steady_momr} the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

@@

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
5\times10^{-3}$. (So the dust layer thickness $\Hd\gg 5
\times10^{-3}\Hgas$ as well.) Then 
instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

>>>>

\subsubsection{Orbital frequency} 
From Eq. \ref{steady_momr} the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio, with $h_\mathrm{g}\ll 1$ for protoplanetary disks.

@@

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas/2\left(1+\epsilon\right)$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
2\times10^{-3}$. This can be met if the dust is not well settled. 
%(So the dust layer thickness $\Hd\gg 2
%\times10^{-3}\Hgas$ as well.) 
Then instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

### 2017-08-04 model.tex
removed higher order correction section for submission, misc edits

In Appendix \ref{var_prin} we derive linearize Eqs. \ref{masseq}, \ref{momeq}, and 
Eq. \ref{eff_energy} to derive the following integral relation: 
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \left(\dd v_z \dd v_r^* + \dd v_z^*\dd v_r\right) B +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,D$ be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

@@

Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_P\ln{\left(1 -
    \tepsilon\right)}$ for the entropy into the standard
expression for said criteria \citep{tassoul78}. This substitution is
consistent with our definition of the effective entropy $\seff$ since
we are considering constant $c_s$.

>>>>

In Appendix \ref{var_prin} we derive linearize Eqs. \ref{masseq}, \ref{momeq}, and 
Eq. \ref{eff_energy} to derive the following integral relation: 
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \left(\dd v_z \dd v_r^* + \dd v_z^*\dd v_r\right) B +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where $^*$ denotes the complex conjugate, 
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,D$ can be 
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

@@

Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} can in fact be obtained by
inserting $C_P\ln{\left(1 - 
    \tepsilon\right)}$ for the entropy into the standard
expression for the Solberg-Hoiland criteria 
\citep{tassoul78} for axisymmetric stability of adiabatic gas. 
This substitution is 
consistent with our definition of the effective entropy $\seff$ since
we are considering constant $c_s$.

### 2017-08-04 paper.tex
removed higher order correction section for submission, misc edits

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follow the gas flow.
  %AY: Edited b/c "This is" was too vague
  This near conservation of dust-to-gas ratio along streamlines
  is analogous to the near conservation of entropy along flows of (dust-free) gas with weak heating and cooling.   
  We develop this thermodynamic analogy into a
  framework to study dusty gas dynamics in protoplanetary disks. We  
  show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this correspondence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilties; 
\item 
  the streaming instability neccessarily involves a gas pressure
  response 
  that lags behind dust density; 
%  the streaming instability neccessarily involves a phase lag between
%  gas pressure and dust densities; 
\item dust-loading introduces buoyancy forces that generally 
  stabilizes the vertical shear 
  instability associated with global radial temperature gradients. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss %future applications of this thermodynamic framework in
%finding 
 dusty analogs of other hydrodynamic 
processes (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), how to simulate dusty protoplanetary disks
with minor tweaks to purely hydrodynamic codes, as well as possible generalizations
of our analogy between dust-laden and nearly adiabatic flows.

@@

This paper is organized as follows. 
In \S\ref{grow_osc} 
we describe a key insight of our study, that pressure (i.e.\ ``PdV") work explains the basis of several instabilities, including SI and VSI, in terms of
pressure-density phase lags.    
We develop our formalism in \S\ref{setup} by transforming the
two fluid equations of a tightly-coupled dusty gas into single fluid
hydrodynamic equations with a special cooling function.  Here we also
define the effective entropy and buoyancy of our model fluid. 
%We derive equilibrium solutions in \S\ref{eqm}, which we use in our
%explicit stability calculations.
In \S\ref{limits} we discuss general stability properties of dusty
disks. We then analyze explicit disk models 
\S\ref{linear_problem}, were we show that dusty edges are unstable,
and revisit the SGI and SI in the tight 
coupling regime. In \S\ref{results} we 
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follow the gas flow.
  %AY: Edited b/c "This is" was too vague
  This near conservation of dust-to-gas ratio along streamlines
  is analogous to the near conservation of entropy along flows of (dust-free) gas with weak heating and cooling.   
  We develop this thermodynamic analogy into a
  framework to study dusty gas dynamics in protoplanetary disks. We  
  show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this correspondence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilties; 
\item 
 radial dust edges are unstable if dust is vertically well-mixed;  
\item 
  the streaming instability neccessarily involves a gas pressure
  response 
  that lags behind dust density; 
%  the streaming instability neccessarily involves a phase lag between
%  gas pressure and dust densities; 
\item dust-loading introduces buoyancy forces that generally 
  stabilizes the vertical shear 
  instability associated with global radial temperature gradients. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss %future applications of this thermodynamic framework in
%finding 
 dusty analogs of other hydrodynamic 
processes (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and how to simulate dusty protoplanetary disks
with minor tweaks to existing codes for pure gas dynamics.%, as well as possible generalizations
%of our analogy between dust-laden and nearly adiabatic flows.

@@

This paper is organized as follows. 
In \S\ref{grow_osc} 
we describe a key insight of our study, that pressure (i.e.\ ``PdV") work explains the basis of several instabilities, including SI and VSI, in terms of
pressure-density phase lags.    
We develop our formalism in \S\ref{setup} by transforming the
two fluid equations of a tightly-coupled dusty gas into single fluid
hydrodynamic equations with a special cooling function.  Here we also
define the effective entropy and buoyancy of our model fluid. 
%We derive equilibrium solutions in \S\ref{eqm}, which we use in our
%explicit stability calculations.
In \S\ref{limits} we discuss general stability properties of dusty
disks. We then analyze explicit disk models 
\S\ref{linear_problem}, were we show that radial dust edges are unstable,
and revisit the SGI and SI in the tight 
coupling regime. In \S\ref{results} we 
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

### 2017-08-04 summary.tex
removed higher order correction section for submission, misc edits

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. We will apply 
our modified code --- d\textsc{Pluto} --- to the streaming instability
and disk-planet interactions in an upcoming publication (Lin et al.,
in preparation).

\subsection{Higher order corrections}
{\bf notes for self. maybe exclude in final. maybe move into appendix}

The simplified one-fluid formulation, 
Eqs. \ref{masseq}---\ref{tempeq}, can 
produce spuriously-growing modes with frequency $|\sigma|\gtrsim
\Omega$ (see Appendix \S\ref{spurious_epi}). This 
results from neglecting terms of $O(|\widetilde{\bm{v}}|^2)$ after converting
the full two-fluid equations to the one-fluid formulation. The full
one-fluid equations 
%is significantly more complex than our 
%model \cite[e.g.][]{laibe14}. It 
 includes an evolutionary equation for $\widetilde{\bm{v}}$, which has no
 hydrodynamic counter-part \citep{laibe14}. However, it is
 still possible to improve upon 
 the first-order formulation whilst retaining similarilty to  
 standard hydrodynamics.  
%without having the complexity of the full equations

The full one-fluid equations of motion includes the acceleration term 
\begin{align}
%\left.\frac{D\bm{v}}{Dt}\right|_\mathrm{add} = 
%\bm{a} =  
-\frac{1}{\rho}\nabla\cdot\left(\frac{\rhod\rhog}{\rho}\widetilde{\bm{v}}\otimes\widetilde{\bm{v}}\right) 
\equiv  \frac{1}{\rho}\nabla\cdot \bm{T} 
\end{align}
\citep{youdin05a}, 
where $\bm{T}$ is an effective stress tensor. 
In the terminal velocity
approximation with $\widetilde{\bm{v}} = \tstop 
\nabla P/\rhog$, we have 
\begin{align}
\bm{T} = - \frac{\epsilon\tstop^2}{\rho}\nabla P \otimes \nabla P.
\end{align}
If we define 
\begin{align}
  \bm{g} \equiv \sqrt{4\pi G \epsilon \rho} \frac{\nabla P}{\rho}
  \tstop,\label{effective_g} 
\end{align}
then
\begin{align}
\bm{T} = - \frac{1}{4\pi G} \bm{g}\otimes\bm{g}, \label{effective_gstress}
\end{align}
%{\bf tensor virial theorems?}
which is just the anisotropic part of the  
gravitational stress tensor in self-gravitating hydrodynamics 
\citep{lynden-bell72}.

 However, unlike self-gravity, where $\bm{g}$ is
determined through the Poisson equation; here $\bm{g}$ is directly
related to $P$ (Eq. \ref{effective_g}), which itself obeys an 
evolutionary equation (Eq. \ref{eff_energy}). 
Nevertheless, Eq. \ref{effective_g}---\ref{effective_gstress}
show that correlations in  pressure gradients lead to 
momentum transport of the mixture as a whole. This is similar to Reynolds stresses in
a turbulent disk \citep{balbus99}. In particular, 
outwards angular momentum transport requires $\p_rP\p_\phi P > 0$.

>>>>

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. We will apply 
this modified code ---d\textsc{Pluto}--- to study dusty disk-planet interaction (Lin et al., 
in preparation). Preliminary results indicate the simplified one-fluid 
approach can reproduce features such as dusty rings associated with 
planet-induced gaps similar to that obtained from explicit two-fluid
simulations \citep[e.g.][]{dong17}.

% \subsection{Higher order corrections}
% {\bf notes for self. maybe exclude in final. maybe move into appendix}

% The simplified one-fluid formulation, 
% Eqs. \ref{masseq}---\ref{tempeq}, can 
% produce spuriously-growing modes with frequency $|\sigma|\gtrsim
% \Omega$ (see Appendix \S\ref{spurious_epi}). This 
% results from neglecting terms of $O(|\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}|^2)$ after converting
% the full two-fluid equations to the one-fluid formulation. The full
% one-fluid equations 
% %is significantly more complex than our 
% %model \cite[e.g.][]{laibe14}. It 
%  includes an evolutionary equation for $\widetilde{\bm{v}}\equiv\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$, which has no
%  hydrodynamic counter-part \citep{laibe14}. However, it is
%  still possible to improve upon 
%  the first-order formulation whilst retaining similarilty to  
%  standard hydrodynamics.  
% %without having the complexity of the full equations

% The full one-fluid equations of motion includes the acceleration term 
% \begin{align}
% %\left.\frac{D\bm{v}}{Dt}\right|_\mathrm{add} = 
% %\bm{a} =  
% -\frac{1}{\rho}\nabla\cdot\left(\frac{\rhod\rhog}{\rho}\widetilde{\bm{v}}\otimes\widetilde{\bm{v}}\right) 
% \equiv  \frac{1}{\rho}\nabla\cdot \bm{T} 
% \end{align}
% \citep{youdin05a}, 
% where $\bm{T}$ is an effective stress tensor. 
% In the terminal velocity
% approximation with $\widetilde{\bm{v}} = \tstop 
% \nabla P/\rhog$, we have 
% \begin{align}
% \bm{T} = - \frac{\epsilon\tstop^2}{\rho}\nabla P \otimes \nabla P.
% \end{align}
% If we define 
% \begin{align}
%   \bm{g} \equiv \sqrt{4\pi G \epsilon \rho} \frac{\nabla P}{\rho}
%   \tstop,\label{effective_g} 
% \end{align}
% then
% \begin{align}
% \bm{T} = - \frac{1}{4\pi G} \bm{g}\otimes\bm{g}, \label{effective_gstress}
% \end{align}
% %{\bf tensor virial theorems?}
% which is just the anisotropic part of the  
% gravitational stress tensor in self-gravitating hydrodynamics 
% \citep{lynden-bell72}.

%  However, unlike self-gravity, where $\bm{g}$ is
% determined through the Poisson equation; here $\bm{g}$ is directly
% related to $P$ (Eq. \ref{effective_g}), which itself obeys an 
% evolutionary equation (Eq. \ref{eff_energy}). 
% Nevertheless, Eq. \ref{effective_g}---\ref{effective_gstress}
% show that correlations in  pressure gradients lead to 
% momentum transport of the mixture as a whole. This is similar to Reynolds stresses in
% a turbulent disk \citep{balbus99}. In particular, 
% outwards angular momentum transport requires $\p_rP\p_\phi P > 0$.
